---
title: "dreval"
author: "Charlotte Soneson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dreval}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: dreval.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    eval = TRUE,
    comment = "#>"
)
```

```{r setup}
library(dreval)
```

# Introduction

The `dreval` package provides functionality to evaluate and compare several
reduced dimension representations, in terms of how well they retain structure
seen in a reference representation. Here, we illustrate its application to a
single-cell RNA-seq data set of peripheral blood mononuclear cells (PBMCs),
representing a subset of the `pbmc3k` data set from the `TENxPBMCData` package.

```{r}
data(pbmc3ksub)
```

# Evaluation

We use the `dreval()` function to calculate several metrics comparing each of
the reduced dimension representations in the `SingleCellExperiment` object to
the underlying normalized and log-transformed data contained in the `logcounts`
slot. For a list of the calculated metrics, see the help page for `dreval()`.

```{r}
dre <- dreval(
  sce = pbmc3ksub, refType = "assay", refAssay = "logcounts", 
  dimReds = NULL, nSamples = 500, kTM = c(10, 75), 
  verbose = FALSE, distNorm = "l2"
)
```

# Result visualization

The `plots` entry of `dre` contains several diagnostic plot objects, while the
`scores` entry contains the calculated scores for each of the evaluated reduced
dimension representations. These can be summarized visually with the
`plotRankSummary()` function.

```{r, fig.width = 8, fig.height = 6}
dre$scores
cowplot::plot_grid(plotlist = dre$plots$disthex)
plotRankSummary(dre$scores)
```

# Using a reduced dimension representation as the reference space

In the example above, we used the original feature space (the `logcounts` assay)
as the reference space, to which all reduced dimension representations were
compared. It is also possible to use one of the reduced dimension
representations as the reference space, as we will illustrate here.

```{r, fig.width = 8, fig.height = 6}
dre <- dreval(
  sce = pbmc3ksub, refType = "dimred", refDimRed = "PCA", 
  dimReds = NULL, nSamples = 500, kTM = c(10, 75), 
  verbose = FALSE, distNorm = "l2"
)
dre$scores
cowplot::plot_grid(plotlist = dre$plots$disthex)
plotRankSummary(dre$scores)
```

# Computational efficiency

`dreval` relies heavily on the calculation of pairwise distances among samples,
and is consequently computationally demanding for large data sets. The
`dreval()` function allows for subsampling a smaller number of samples, as well
as a selection of a subset of the variables, to speed up calculations and reduce
the memory footprint. Here we investigate the stability of the calculated
metrics as decreasing numbers of samples are selected.

```{r, fig.width = 10, fig.height = 8, eval = FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})

set.seed(1)
nS <- c(100, 400, 1000, 1500, 2700)
v <- lapply(nS, function(n) {
  if (n == 2700) nR <- 1
  else nR <- 2
  lapply(1:nR, function(i) {
    message(n, " samples, replicate ", i)
    dreval(sce = pbmc3ksub, refAssay = "logcounts", dimReds = NULL,
           nSamples = n, kTM = 50, verbose = FALSE)$scores %>%
      dplyr::mutate(nSamples = n, replicate = i)
  })
})
vv <- do.call(dplyr::bind_rows, v)

vv <- vv %>%
  tidyr::gather(
    key = "measure", value = "value", 
    -Method, -dimensionality, -nSamples, -replicate
  )


suppressWarnings(
  ggplot(vv, aes(x = nSamples, y = value, color = Method)) + 
    geom_point(size = 3) + facet_wrap(~ measure, scales = "free_y") + 
    geom_smooth(se = FALSE) + theme_bw()
)
```

# Session info

```{r}
sessionInfo()
```

# References
